fairmetrics: An R package for group fairness evaluation

### 2025-09-11 utils/joss_paper/paper.tex
small fix 'Department of Statistical Sciences'
fixed references

      \affil[1]{Department of Statistical Science, University of
Toronto}
  \date{\vspace{-5ex}}

@@

Fairness is a growing area of machine learning (ML) that focuses on
ensuring that models do not produce systematically biased outcomes
across groups defined by protected attributes, such as race, gender, or
age. The \texttt{fairmetrics} R package provides a user-friendly
framework for rigorously evaluating group-based fairness criteria,
including independence (e.g., statistical parity), separation (e.g.,
equalized odds), and sufficiency (e.g., predictive parity) for binary
protected attributes. The package provides both point and interval
estimates for a variety of commonly used criteria. \texttt{fairmetrics}
also includes an example dataset derived from the Medical Information
Mart for Intensive Care, version II (MIMIC-II) database (Goldberger et
al. 2000; J. Raffa 2016) to demonstrate its use.

@@

\NormalTok{           Fairness Assesment                                  Metric}
\DecValTok{1}\NormalTok{          Statistical Parity                Positive Prediction Rate}
\DecValTok{2}\NormalTok{           Equal Opportunity                     False Negative Rate}
\DecValTok{3}\NormalTok{         Predictive Equality                     False Positive Rate}
\DecValTok{4}\NormalTok{  Balance }\ControlFlowTok{for}\NormalTok{ Positive Class           Avg. Predicted Positive Prob.}
\DecValTok{5}\NormalTok{  Balance }\ControlFlowTok{for}\NormalTok{ Negative Class           Avg. Predicted Negative Prob.}
\DecValTok{6}\NormalTok{  Positive Predictive Parity               Positive Predictive Value}
\DecValTok{7}\NormalTok{  Negative Predictive Parity               Negative Predictive Value}
\DecValTok{8}\NormalTok{          Brier Score Parity                             Brier Score}
\DecValTok{9}\NormalTok{     Overall Accuracy Parity                                Accuracy}
\DecValTok{10}\NormalTok{         Treatment }\FunctionTok{Equality}\NormalTok{ (False Negative)}\SpecialCharTok{/}\NormalTok{(False Positive) Ratio}

@@

\bibitem[\citeproctext]{ref-fairness_package}
Kozodoi, Nikita, and Tibor V. Varga. 2021. \emph{Fairness: Algorithmic
Fairness Metrics}. \url{https://doi.org/10.32614/cran.package.fairness}.

\bibitem[\citeproctext]{ref-mattuMachineBias}
Mattu, Lauren Kirchner, Jeff Larson. 2016. {``Machine {Bias}.''}
\emph{ProPublica}.
https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing.

@@

\bibitem[\citeproctext]{ref-mlr3fairness_package}
Pfisterer, Florian, Wei Siyi, and Michel Lang. 2024. \emph{Mlr3fairness:
Fairness Auditing and Debiasing for 'Mlr3'}.
\url{https://doi.org/10.32614/cran.package.mlr3fairness}.

\bibitem[\citeproctext]{ref-raffa2016clinical}
Raffa, Jesse. 2016. {``Clinical Data from the MIMIC-II Database for a
Case Study on Indwelling Arterial Catheters (Version 1.0).''}
\url{https://doi.org/10.13026/C2NC7F}.
\url{https://doi.org/10.13026/C2NC7F}.

@@

\bibitem[\citeproctext]{ref-wisniewski2022fairmodels}
Wiśniewski, Jakub, and Przemysław Biecek. 2022. {``Fairmodels: A
Flexible Tool for Bias Detection, Visualization, and Mitigation in
Binary Classification Models.''} \emph{The R Journal} 14 (1): 227--43.
\url{https://doi.org/10.32614/RJ-2022-019}.

>>>>

      \affil[1]{Department of Statistical Sciences, University of
Toronto}
  \date{\vspace{-5ex}}

@@

Fairness is a growing area of machine learning (ML) that focuses on
ensuring that models do not produce systematically biased outcomes
across groups defined by protected attributes, such as race, gender, or
age. The \texttt{fairmetrics} R package provides a user-friendly
framework for rigorously evaluating group-based fairness criteria,
including independence (e.g., statistical parity), separation (e.g.,
equalized odds), and sufficiency (e.g., predictive parity) for binary
protected attributes. The package provides both point and interval
estimates for a variety of commonly used criteria. \texttt{fairmetrics}
also includes an example dataset derived from the Medical Information
Mart for Intensive Care, version II (MIMIC-II) database (Goldberger et
al. 2000; J. Raffa 2016; J. D. Raffa et al. 2016) to demonstrate its
use.

@@

\NormalTok{           Fairness Assessment                                 Metric}
\DecValTok{1}\NormalTok{          Statistical Parity                Positive Prediction Rate}
\DecValTok{2}\NormalTok{           Equal Opportunity                     False Negative Rate}
\DecValTok{3}\NormalTok{         Predictive Equality                     False Positive Rate}
\DecValTok{4}\NormalTok{  Balance }\ControlFlowTok{for}\NormalTok{ Positive Class           Avg. Predicted Positive Prob.}
\DecValTok{5}\NormalTok{  Balance }\ControlFlowTok{for}\NormalTok{ Negative Class           Avg. Predicted Negative Prob.}
\DecValTok{6}\NormalTok{  Positive Predictive Parity               Positive Predictive Value}
\DecValTok{7}\NormalTok{  Negative Predictive Parity               Negative Predictive Value}
\DecValTok{8}\NormalTok{          Brier Score Parity                             Brier Score}
\DecValTok{9}\NormalTok{     Overall Accuracy Parity                                Accuracy}
\DecValTok{10}\NormalTok{         Treatment }\FunctionTok{Equality}\NormalTok{ (False Negative)}\SpecialCharTok{/}\NormalTok{(False Positive) Ratio}

@@

\bibitem[\citeproctext]{ref-fairness_package}
Kozodoi, Nikita, and Tibor V. Varga. 2021. \emph{{fairness}: Algorithmic
Fairness Metrics}. \url{https://doi.org/10.32614/cran.package.fairness}.

\bibitem[\citeproctext]{ref-mattuMachineBias}
Mattu, Lauren Kirchner, Jeff Larson. 2016. {``Machine {Bias}.''}
\emph{ProPublica}.
\url{https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing}.

@@

\bibitem[\citeproctext]{ref-mlr3fairness_package}
Pfisterer, Florian, Wei Siyi, and Michel Lang. 2024.
\emph{{mlr3fairness}: Fairness Auditing and Debiasing for 'Mlr3'}.
\url{https://doi.org/10.32614/cran.package.mlr3fairness}.

\bibitem[\citeproctext]{ref-raffa2016clinical}
Raffa, Jesse. 2016. {``Clinical Data from the MIMIC-II Database for a
Case Study on Indwelling Arterial Catheters (Version 1.0).''}
\url{https://doi.org/10.13026/C2NC7F}.

@@

\bibitem[\citeproctext]{ref-wisniewski2022fairmodels}
Wiśniewski, Jakub, and Przemysław Biecek. 2022. {``{fairmodels}: A
Flexible Tool for Bias Detection, Visualization, and Mitigation in
Binary Classification Models.''} \emph{The R Journal} 14 (1): 227--43.
\url{https://doi.org/10.32614/RJ-2022-019}.
